Solving the independent set problem by generic programming tensor networks \thanks{\funding{...}}

### 2021-08-12 paper/paper.tex
update

\title{Solving the independent set problem by generic programming einsum networks
\thanks{\funding{...}}
}

@@

\begin{abstract}
	This paper is about solving the independent set problem by eincoding this problem to an einsum network. 
    We show how to obtain the maximum independent set size, the independence polynomial and optimal configurations of a graph by engineering the tensor element algebra.
    We also show how to analyse the local properties of a graph by contracting an open einsum network.
\end{abstract}

% REQUIRED
\begin{keywords}
  maximum independent set, einsum network
\end{keywords}

@@

\section{Introduction}
In this work, we introduce a tensor based framework to study the famous graph problem of finding independent sets.
Given an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set that for any $u,v \in I$, there is no edge connecting $u$ and $v$ in $G$.
The problem of finding the maximum independent set (MIS) size $\alpha(G) \equiv \max_{I}|I|$ belongs to the complexity class NP-complete~\cite{Hastad1996}, which is unlikely to be decided in polynomial time.
It is hard to even approximate this size in polynomial time within a factor $|V|^{1-\epsilon}$ for an arbitrarily small positive $\epsilon$.
The exhaustive search for a solution costs time $2^{|V|}$.
More efficient algorithms to compute the MIS size exactly includes the branching algorithm and dynamic programming.
Without changing the fact of exponential scaling in computing time, the branching algorithm gives a smaller base.
For example, in ~\cite{Xiao2017}, a sophisticated branching algorithm has a time complexity $1.1893^n n ^{O(1)}$.
The dynamic programming approach~\cite{Courcelle1990,Fomin2013} works better for graphs with small tree width $tw(G)$, it gives an algorithms of complexity $O(2^{tw(G)}tw(G)n)$.
People are interested in solving the independent set problem better not only because it is an NP-complete problem that directly related to other NP-complete problems like maximal cliques and vertex cover~\cite{Moore2011},
but also for its close relation with physical applications like hard spheres lattice gas model~\cite{Dyre2016}, and Rydberg hamiltonian~\cite{Pichler2018}.
However, in these applications, knowing the MIS size and one of the optimum solutions is not the only goal.
%A set is independent if and only if it is a clique in the graph's complement, so the two concepts are complementary.
%A set is independent if and only if its complement is a vertex cover. Therefore, the sum of the size of the largest independent set $\alpha (G)$ and the size of a minimum vertex cover $\beta (G)$ is equal to the number of vertices in the graph.
People often ask different questions about independent sets in order to understand the landscape of their models better.
These questions includes but not limited to, counting all independent sets, obtaining all independent sets of size $\alpha(G)$ and $\alpha(G)-1$,
counting the number of (maximal) independent sets of different sizes, and understanding the effect of a local gadget.
In this work, we attack this problem by mapping it to an generic ``einsum'' network.
It does not give a better time complexity comparing to dynamic programming, but is versatile enough to answer the above questions by engineering the tensor elements with minimum effort.

\section{Einsum networks}
%The word ``einsum'' is a shorthand for Einstein's summation, however, modern einsum notation in program is actually invented by a group of programmers.
Einstein's notation is originally proposed as a generalization to of binary matrix multiplication to n-ary tensor contraction.
Let $A, B$ be two matrices, the matrix multiplication is defined as $C_{ik} = \sum_{j}A_{ij}B_{jk}$.
In Einstein's notation, it is denoted as $C_i^k = A_i^j B_j^k$, where the paired subscript and superscript $j$ is a dummy index summed over, hence each index appears precisely twice.
When we have multiple tensors doing the above sum-product operation, we get a tensor network~\cite{Orus2014}.
A tensor network can be represented as a mutigraph with open edges by viewing a tensor as a vertex, a label pairing two tensors as an edge, and the remaining unpaired labels as open edges.

\begin{example}
    A tensor networks $C_i^l = A_{ij}^{km} B^l_{km} V^j$ has the following multigraph representation.

@@

Einsum network is a generalization of tensor network by not restricting the number of times a label appears in the notation,
hence whether an index is a superscript or a subscript makes no sense now.
It is also called sum-product network or factor graph~\cite{Bishop2006} in some contexts.
The graphical representation of an einsum is a hypergraph, where an edge (label) can be shared by an arbitrary number of vertices (tensors).

\begin{example}
$C_{ijk} = A_{jkm} B_{mil} V_{jm}$ is an einsum network, it represents $C_{ijk} = \sum_{ml}A_{jkm} B_{mia} V_{jm}$.
Its hypergraph representation is as the following, where we use different color to annotate different hyperedges.

@@

In the main text, we stick to the einsum notation rather than the tensor network notation.
As a note to those who are more familiar with tensor network representation,
although one can easily translate an einsum network to the equivalent tensor network by adding $\delta$ tensors (a generalization of identity matrix to higher order).
It can sometime increase the contraction complexity of a graph. We have an example demonstrating this point in \App{app:tensorbad}.

\section{Independence polynomial}
One can encode the independence polynomial~\cite{Ferrin2014, Harvey2017} of $G$ to an einsum network.
Independence polynomial is an important graph polynomial that contains the counting information of an independent set problem. It is defined as
\begin{equation}
I(G, x) = \sum_{k=1}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in $G$, and $\alpha(G)$ is the maximum independent set size.
We encode this polynomial to an einsum network by placing a rank one tensor of size $2$ parametrized by $x_i$ on a vertex $i$
\begin{equation}
    W(x_i)_{s_i} = \left(\begin{matrix}
        1 \\
        x_i
    \end{matrix}\right)_{s_i},\label{eq:wtensor}
\end{equation}
and a rank two tensor of size $2 \times 2$ on an edge $(i,j)$
\begin{equation}
    B_{s_i s_j} = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right)_{s_is_j},\label{eq:btensor}
\end{equation}
where a tensor index $s_i$ is a boolean variable that having the meaning of being 1 if vertex $i$ is in the independent set, 0 otherwise.
It corresponds to a hyperedge in the hypergraph.
The contraction of such an einsum network gives
\begin{equation}
    P(G, \{x_1,\ldots,x_n\}) = \sum\limits_{s_1, s_2, \ldots, s_n = 0}^{1} \prod\limits_{i=1}^n W(x_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j},
\end{equation}
where the summation runs over all vertex configurations $\{s_1,\ldots,s_n\}$ and accumulates the product of tensor elements to the scalar output $P$.
We can see an edge tensor represents the restriction on an edge that if both vertices connected by it are included in the set, then such configuration has no contribution to the output.
When we set $x_i = x$, the contraction result corresponds to the independence polynomial.
One can see the connection from the fact that the product over vertex tensor elements gives a factor $x^k$, where $k=\sum_i s_i$ counts the set size, 
and the product over edge tensor elements gives a factor $1$ for a configuration being in an independent set, $0$ otherwise.
One directly benefit of mapping the independent set problem to an einsum network is one can take the advantage of recently developed techniques in tensor network based quantum circuit simulations ~\cite{Gray2021,Pan2021},
where people evaluate an einsum network by pairwise contracting tensors in a heuristic order.
A good contraction order can reduce the time complexity significantly, at the cost of having a space overhead of $O(2^{tw(G)})$.
Here $tw(G)$ is the tree width of the line graph of an einsum hypergraph, while the line graph of an einsum hypergraph corresponds to the original graph $G$ that we mapped from.~\cite{Markov2008}
The pairwise tensor contraction also makes it possible to utilize basic linear algebra subprograms (BLAS) functions to speed up our computation for certain tensor element types.

\begin{example}
    Mapping a graph (left) to an einsum network, the resulting einsum network is shown in the right panel.
    In the einsum's graphical representation, a vertex is mapped to a hyperedge,
    and an edge is mapped to an edge tensor.

@@

Before contracting the einsum network and evaluating the independence polynomial numerically, let us first give up thinking $0$s and $1$s in tensors $W(x)$ and $B$ as regular computer numbers such as integers and floating point numbers.
Instead, we treat them as the additive identity and multiplicative identity of a commutative semiring.
A semiring is a ring without additive inverse, while a commutative semiring is a semiring that multiplication commutative.
To define a commutative semiring with addition algebra $\oplus$ and multiplication algebra $\odot$ on a set $R$, the following relations must hold for arbitrary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b + a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}

In the following, we show how to obtain the independence polynomial, the maximum independent set size and optimal configurations of a general graph $G$ by designing tensor element types as commutative semirings,
i.e. making the einsum network generic~\cite{Stepanov2014}.

\subsection{The polynomial approach}
A straight forward approach to evaluate the independence polynomial is treating the tensor elements as polynomials, and evaluate the polynomial directly.
Let us create a polynomial type, and represent a polynomial $a_0 + a_1 x + \ldots + a_k x^k$ as a vector $(a_0, a_1, \ldots, a_k) \in R^k$, e.g. $x$ is represented as $(0, 1)$.
We define the algebra between the polynomials $a$ of order $k_a$ and $b$ of order $k_b$ as
\begin{align}
    \begin{split}
    a \oplus b &= (a_0 + b_0, a_1 + b_1, \ldots, a_{\max(k_a, k_b)} + b_{\max(k_a, k_b)}),\\
    a \odot b &= (a_0 + b_0, a_1b_0 + a_0b_1, \ldots, a_{k_a} b_{k_b}),\\
    \mymathbb{0} &= (),\\
    \mymathbb{1} &= (1).\label{eq:polynomial}
    \end{split}
\end{align}
By contracting the einsum network with polynomial type, the final result is the exact representation of the independence polynomial.
In the program, the multiplication can be evaluated efficiently with the convolution theorem~\cite{Schonhage1971}.
The only problem of this method is it suffers from a space overhead that proportional to the maximum independent set size because each polynomial requires a vector of such size to store the factors.
In the following subsections, we managed to solve this problem.

\subsection{The fitting and Fourier transformation approaches}
Let $m=\alpha(G)$ be the maximum independent set size and $X$ be a set of real numbers of cardinality $m+1$.
We compute the einsum contraction for each $x_i \in X$ and obtain the following relations
\begin{align}
    \begin{split}
a_0 + a_1 x_1 + a_1 x_1^2 + \ldots + a_m x_1^m &= y_0\\
a_0 + a_1 x_2 + a_2 x_2^2 + \ldots + a_m x_2^m &= y_1\\
\ldots&\\
a_0 + a_1 x_m + a_2 x_m^2 + \ldots + a_m x_m^m& = y_m
    \end{split}
\end{align}
The polynomial fitting between $X$ and $Y = \{y_0, y_1, \ldots, y_m\}$ gives us the factors.
The polynomial fitting is essentially about solving the following linear equation
\begin{align}
\left(\begin{matrix}
1 & x_1 & x_1^2 & \ldots & x_1^m \\
1 & x_2 & x_2^2 & \ldots & x_2^m \\
\vdots & \vdots & \vdots &\ddots & \vdots \\
1 & x_m & x_m^2 & \ldots & x_m^m
\end{matrix}\right)
\left(\begin{matrix}
a_0 \\ a_1 \\ \vdots \\ a_m
\end{matrix}\right)
= \left(\begin{matrix}
y_0 \\ y_1 \\ \vdots \\ y_m
\end{matrix}\right).\label{eq:lineareq}
\end{align}

@@

\subsection{Maximal independence polynomial}
Some times people are interested in knowing maximal solutions to understand why their program is trapped in a local minimal.
Then they might want to compute the maximal independence polynomial.
Let us denote the neighbour of a vertex $v$ as $N(v)$ and $N[v] = N(v)\cup \{v\}$.
A maximal independent set $I_m$ is an independent sets that there does not exist a vertex $v$ that $N[v] \cap I_m = \emptyset$.
Let us modify the einsum network for computing independence polynomial by adding this restriction.
Instead of defining the restriction on vertices and edges, we define it on $N[v]$
\begin{align}
    T(x_v)_{s_1,s_2,\ldots,s_{|N(v)|},s_v} = \begin{cases}
        s_vx_v & s_1=s_2=\ldots=s_{|N(v)|}=0,\\
        1-s_v& otherwise.\\
    \end{cases}
\end{align}
As an example, for a vertex of degree 2, the resulting rank 3 tensor is
\begin{align}
    T(x_v)=\left(\begin{matrix}
    \left(\begin{matrix}
        0 &1 \\
        1 &1
    \end{matrix}\right)\\
    \left(\begin{matrix}
        x_v &0 \\
        0 &0
    \end{matrix}\right)
    \end{matrix}\right).
\end{align}

We do the same computation as independence polynomial, the coefficients of resulting polynomial gives the counting of maximal independent sets, or the maximal independence polynomial.
The computational complexity of this new einsum network is often larger than the one for computing independence polynomial.
However, in many sparse graphs, this einsum network contraction approach is still much faster than computing the maximal cliques on its complement by applying the Bron-Kerbosch algorithm.

@@

\section{Enumerating configurations}
One may also want to obtain all solutions, it can be achieved replacing the factors $a_x$ with a set of bit strings $s_x$,
We design a new element type having the following algebra
\begin{align}
\begin{split}
    s \oplus t &= s \cup t\\
    s \odot t &= \{\sigma \lor^\circ \tau | \sigma \in s, \tau \in t\}\\
    \mymathbb{0} &= \{\}\\
    \mymathbb{1} &= \{0^{\otimes n}\}
\end{split}\label{eq:set}
\end{align}
\iffalse
\begin{align}
\begin{split}
    s_x\infty^x \oplus s_y\infty^y &= \begin{cases}
        (s_x \cup s_y)\infty^{\max(x,y)}, & x = y\\
        s_y\infty^{\max(x,y)}, & x < y\\
        s_x\infty^{\max(x,y)}, & x > y
    \end{cases},\\
    s_x\infty^x \odot s_y\infty^y &= \{\sigma \lor^\circ \tau | \sigma \in s_x, \tau \in s_y\}\infty^{x+y},\\
    \mymathbb{0} &= \{\}\infty^{-\infty},\\
    \mymathbb{1} &= \{0^{\otimes n}\}\infty^{0},
\end{split}
\end{align}
\fi
where $\lor^\circ$ is the Hadamard logic or operation over two bit strings, which means joining of two local configurations.
The variable $x$ in the vertex tensor is initialized to $x_i = \{\boldsymbol{e}_{i}\}$,
where $\boldsymbol{e}_i$ is a one hot vector of size $|G|$.
One can easily check this algebra is a commutative semiring.
When we use the above algebra as factors of independence polynomials, the resulting algebra is also a commutative semiring.
With this new element type, the einsum network contraction will give all solutions rather than just a number for counting.
By slightly modifying the above algebra, it can also be used to obtain just a single configuration to save the computational effort.
%We leave this as an exercise for readers.

@@

\subsection{Bounding the enumeration space}
When one uses the set algebra in \Eq{eq:set} to represent the factors in \Eq{eq:countingtropical} for enumerating all optimum configurations, he will find the program stores more than necessary intermediate configurations and cause significant overheads in space.
To speed up the computation, we use $\alpha(G)$ to bound the searching space.
We first compute the value of $\alpha(G)$ with tropical numbers and cache all intermediate tensors.
Then we compute a boolean masks for each cached tensor, where we use a boolean true to represent a tensor element having contribution to the maximum independent set (i.e. with a non-zero gradient) and boolean false otherwise.
Finally, we perform masked matrix multiplication using the new element type with the above algebra for obtaining all configurations.
Notice that these masks are in fact tensor elements with non-zero gradients with respect to MIS size, we compute these masks by back propagating gradients.
To derive the backward rule, we consider a tropical matrix multiplication $C = A B$, we have the following inequality
\begin{align}
    A_{ij} \odot B_{jk} &\leq C_{ik}.
\end{align}
Moving $B_{ik}$ to the right hand side, we have
\begin{align}
    A_{ij} &\leq (\oplus_{k} (C_{ik}^{-1} \odot B_{jk}))^{-1}
\end{align}
where the tropical multiplicative inverse is defined as the additive inverse of the regular algebra. The equality holds if and only if element $A_{ij}$ contributions to $C$ (i.e. has non-zero gradient).
Let the mask for $C$ being $\overline C$, the backward rule for ``gradient'' masks reads
\begin{align}
\overline{A}_{ij} = \delta(A_{ij}, ((C^{\circ-1} \circ \overline C )B^T)_{ij}^{\circ -1}),
\end{align}
where ${}^{\circ -1}$ is the Hadamard inverse, $\circ$ is the Hadamard product, boolean false is treated as tropical zero and boolean true is treated as tropical one.
This rule defined on matrix multiplication can be easily generalized to einsum by replacing the matrix multiplication between $C^{\circ-1} \circ \overline C$ and $B^T$ by an einsum operation.~\cite{} \blue{maybe add an appendix?}  % my blog

>>>>

\title{Solving the independent set problem by generic programming tensor networks
\thanks{\funding{...}}
}

@@

\begin{abstract}
	This paper is about solving the independent set problem by eincoding this problem to a tensor network. 
    We show how to obtain the maximum independent set size, the independence polynomial and optimal configurations of a graph by engineering the tensor element algebra.
    We also show how to analyse the local properties of a graph by contracting an open tensor network.
\end{abstract}

% REQUIRED
\begin{keywords}
  maximum independent set, tensor network
\end{keywords}

@@

\section{Introduction}
In this work, we introduce a tensor based framework to study the famous graph problem of finding independent sets.
Given an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set that for any $u,v \in I$, there is no edge connecting $u$ and $v$ in $G$.
The problem of finding the maximum independent set (MIS) size $\alpha(G) \equiv \max_{I}|I|$ belongs to the complexity class NP-complete~\cite{Hastad1996}, which is unlikely to be decided in polynomial time.
It is hard to even approximate this size in polynomial time within a factor $|V|^{1-\epsilon}$ for an arbitrarily small positive $\epsilon$.
The exhaustive search for a solution costs time $2^{|V|}$.
More efficient algorithms to compute the MIS size exactly includes the branching algorithm and dynamic programming.
Without changing the fact of exponential scaling in computing time, the branching algorithm gives a smaller base.
For example, in ~\cite{Xiao2017}, a sophisticated branching algorithm has a time complexity $1.1893^n n ^{O(1)}$.
The dynamic programming approach~\cite{Courcelle1990,Fomin2013} works better for graphs with small tree width $tw(G)$, it gives an algorithms of complexity $O(2^{tw(G)}tw(G)n)$.
People are interested in solving the independent set problem better not only because it is an NP-complete problem that directly related to other NP-complete problems like maximal cliques and vertex cover~\cite{Moore2011},
but also for its close relation with physical applications like hard spheres lattice gas model~\cite{Dyre2016}, and Rydberg hamiltonian~\cite{Pichler2018}.
However, in these applications, knowing the MIS size and one of the optimum solutions is not the only goal.
%A set is independent if and only if it is a clique in the graph's complement, so the two concepts are complementary.
%A set is independent if and only if its complement is a vertex cover. Therefore, the sum of the size of the largest independent set $\alpha (G)$ and the size of a minimum vertex cover $\beta (G)$ is equal to the number of vertices in the graph.
People often ask different questions about independent sets in order to understand the landscape of their models better.
These questions includes but not limited to, counting all independent sets, obtaining all independent sets of size $\alpha(G)$ and $\alpha(G)-1$,
counting the number of (maximal) independent sets of different sizes, and understanding the effect of a local gadget.
In this work, we attack this problem by mapping it to an generic tensor network.
It does not give a better time complexity comparing to dynamic programming, but is versatile enough to answer the above questions by engineering the tensor elements with minimum effort.

\section{Tensor networks}
%The word ``einsum'' is a shorthand for Einstein's summation, however, modern einsum notation in program is actually invented by a group of programmers.
A tensor network can be viewed as a generalization to of binary matrix multiplication to n-ary tensor contraction.
Let $A, B$ be two matrices, the matrix multiplication is defined as $C_{ik} = \sum_{j}A_{ij}B_{jk}$.
A traditional tensor network refers to the Einstein's notation. In this notation, the matrix multiplication is denoted as $C_i^k = A_i^j B_j^k$, where the paired subscript and superscript $j$ is a dummy index summed over, hence each index appears precisely twice.
When we have multiple tensors doing the above sum-product operation, we get a traditional tensor network~\cite{Orus2014}.
A traditional tensor network can be represented as a mutigraph with open edges by viewing a tensor as a vertex, a label pairing two tensors as an edge, and the remaining unpaired labels as open edges.

\begin{example}
   A traditional tensor network $C_i^l = A_{ij}^{km} B^l_{km} V^j$ has the following multigraph representation.

@@

Here, we want to use a generalized tensor network notation by not restricting the number of times a label appears in the notation,
hence whether an index is a superscript or a subscript makes no sense now.
It is also called einsum, sum-product network or factor graph~\cite{Bishop2006} in some contexts.
The graphical representation of a tensor network in this paper is a hypergraph, where an edge (label) can be shared by an arbitrary number of vertices (tensors).

\begin{example}
$C_{ijk} = A_{jkm} B_{mil} V_{jm}$ is a tensor network, it represents $C_{ijk} = \sum_{ml}A_{jkm} B_{mia} V_{jm}$.
Its hypergraph representation is as the following, where we use different color to annotate different hyperedges.

@@

In the main text, we stick to the our generalized tensor network notation rather than the traditional notation.
As a note to those who are more familiar with the traditional tensor network representation,
although one can easily translate a generalized tensor network to the equivalent traditional tensor network by adding $\delta$ tensors (a generalization of identity matrix to higher order).
It can sometime increase the contraction complexity of a graph. We have an example demonstrating this point in \App{app:tensorbad}.

\section{Independence polynomial}
One can encode the independence polynomial~\cite{Ferrin2014, Harvey2017} of $G$ to a tensor network.
Independence polynomial is an important graph polynomial that contains the counting information of an independent set problem. It is defined as
\begin{equation}
I(G, x) = \sum_{k=1}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in $G$, and $\alpha(G)$ is the maximum independent set size.
The problem of computing independence polynomial belongs to the complexity class \#P-hard.
A traditional approach to compute independence polynomial rigorusly requires a computing time $O(1.442^n)$~\cite{Ferrin2014}\blue{I am not sure about this complexity, this is baed on the naive analysis of theorem 2.2 in ~\cite{Ferrin2014}}.
%with which computing the independence polynomial for a square lattice of size $30\times 30$ would be impossible.
There are some interests in approximating this polynomial efficiently~\cite{Harvey2018}, but here, we focus on the rigorous approaches.
We encode this polynomial to a tensor network by placing a rank one tensor of size $2$ parametrized by $x_i$ on a vertex $i$
\begin{equation}
    W(x_i)_{s_i} = \left(\begin{matrix}
        1 \\
        x_i
    \end{matrix}\right)_{s_i},\label{eq:wtensor}
\end{equation}
and a rank two tensor of size $2 \times 2$ on an edge $(i,j)$
\begin{equation}
    B_{s_i s_j} = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right)_{s_is_j},\label{eq:btensor}
\end{equation}
where a tensor index $s_i$ is a boolean variable that having the meaning of being 1 if vertex $i$ is in the independent set, 0 otherwise.
It corresponds to a hyperedge in the hypergraph.
The contraction of such a tensor network gives
\begin{equation}
    P(G, \{x_1,\ldots,x_n\}) = \sum\limits_{s_1, s_2, \ldots, s_n = 0}^{1} \prod\limits_{i=1}^n W(x_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j},
\end{equation}
where the summation runs over all vertex configurations $\{s_1,\ldots,s_n\}$ and accumulates the product of tensor elements to the scalar output $P$.
We can see an edge tensor represents the restriction on an edge that if both vertices connected by it are included in the set, then such configuration has no contribution to the output.
When we set $x_i = x$, the contraction result corresponds to the independence polynomial.
One can see the connection from the fact that the product over vertex tensor elements gives a factor $x^k$, where $k=\sum_i s_i$ counts the set size, 
and the product over edge tensor elements gives a factor $1$ for a configuration being in an independent set, $0$ otherwise.
One directly benefit of mapping the independent set problem to a tensor network is one can take the advantage of recently developed techniques in tensor network based quantum circuit simulations ~\cite{Gray2021,Pan2021},
where people evaluate a tensor network by pairwise contracting tensors in a heuristic order.
A good contraction order can reduce the time complexity significantly, at the cost of having a space overhead of $O(2^{tw(G)})$.
Here $tw(G)$ is the tree width of the line graph of a tensor network hypergraph, while the line graph of a tensor network hypergraph corresponds to the original graph $G$ that we mapped from.~\cite{Markov2008}
The pairwise tensor contraction also makes it possible to utilize basic linear algebra subprograms (BLAS) functions to speed up our computation for certain tensor element types.

\begin{example}
    Mapping a graph (left) to a tensor network, the resulting tensor network is shown in the right panel.
    In the generalize tensor network's graphical representation, a vertex is mapped to a hyperedge,
    and an edge is mapped to an edge tensor.

@@

Before contracting the tensor network and evaluating the independence polynomial numerically, let us first give up thinking $0$s and $1$s in tensors $W(x)$ and $B$ as regular computer numbers such as integers and floating point numbers.
Instead, we treat them as the additive identity and multiplicative identity of a commutative semiring.
A semiring is a ring without additive inverse, while a commutative semiring is a semiring that multiplication commutative.
To define a commutative semiring with addition algebra $\oplus$ and multiplication algebra $\odot$ on a set $R$, the following relations must hold for arbitrary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b + a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}

In the following, we show how to obtain the independence polynomial, the maximum independent set size and optimal configurations of a general graph $G$ by designing tensor element types as commutative semirings,
i.e. making the tensor network generic~\cite{Stepanov2014}.

\subsection{The polynomial approach}
A straight forward approach to evaluate the independence polynomial is treating the tensor elements as polynomials, and evaluate the polynomial directly.
Let us create a polynomial type, and represent a polynomial $a_0 + a_1 x + \ldots + a_k x^k$ as a vector $(a_0, a_1, \ldots, a_k) \in R^k$, e.g. $x$ is represented as $(0, 1)$.
We define the algebra between the polynomials $a$ of order $k_a$ and $b$ of order $k_b$ as
\begin{align}
    \begin{split}
    a \oplus b &= (a_0 + b_0, a_1 + b_1, \ldots, a_{\max(k_a, k_b)} + b_{\max(k_a, k_b)}),\\
    a \odot b &= (a_0 + b_0, a_1b_0 + a_0b_1, \ldots, a_{k_a} b_{k_b}),\\
    \mymathbb{0} &= (),\\
    \mymathbb{1} &= (1).\label{eq:polynomial}
    \end{split}
\end{align}
By contracting the tensor network with polynomial type, the final result is the exact representation of the independence polynomial.
In the program, the multiplication can be evaluated efficiently with the convolution theorem~\cite{Schonhage1971}.
The only problem of this method is it suffers from a space overhead that proportional to the maximum independent set size because each polynomial requires a vector of such size to store the factors.
In the following subsections, we managed to solve this problem.

\subsection{The fitting and Fourier transformation approaches}
Let $m=\alpha(G)$ be the maximum independent set size and $X$ be a set of real numbers of cardinality $m+1$.
We compute the tensor network contraction for each $x_i \in X$ and obtain the following relations
\begin{align}
    \begin{split}
a_0 + a_1 x_1 + a_1 x_1^2 + \ldots + a_m x_1^m &= y_0\\
a_0 + a_1 x_2 + a_2 x_2^2 + \ldots + a_m x_2^m &= y_1\\
\ldots&\\
a_0 + a_1 x_m + a_2 x_m^2 + \ldots + a_m x_m^m& = y_m
    \end{split}
\end{align}
The polynomial fitting between $X$ and $Y = \{y_0, y_1, \ldots, y_m\}$ gives us the factors.
The polynomial fitting is essentially about solving the following linear equation
\begin{align}
\left(\begin{matrix}
1 & x_1 & x_1^2 & \ldots & x_1^m \\
1 & x_2 & x_2^2 & \ldots & x_2^m \\
\vdots & \vdots & \vdots &\ddots & \vdots \\
1 & x_m & x_m^2 & \ldots & x_m^m
\end{matrix}\right)
\left(\begin{matrix}
a_0 \\ a_1 \\ \vdots \\ a_m
\end{matrix}\right)
= \left(\begin{matrix}
y_0 \\ y_1 \\ \vdots \\ y_m
\end{matrix}\right).\label{eq:lineareq}
\end{align}

@@

\subsection{Maximal independence polynomial}
Some times people are interested in knowing maximal solutions to understand why their program is trapped in a local minimal.
Then they might want to compute the maximal independence polynomial.
Let us denote the neighbour of a vertex $v$ as $N(v)$ and $N[v] = N(v)\cup \{v\}$.
A maximal independent set $I_m$ is an independent sets that there does not exist a vertex $v$ that $N[v] \cap I_m = \emptyset$.
Let us modify the tensor network for computing independence polynomial by adding this restriction.
Instead of defining the restriction on vertices and edges, we define it on $N[v]$
\begin{align}
    T(x_v)_{s_1,s_2,\ldots,s_{|N(v)|},s_v} = \begin{cases}
        s_vx_v & s_1=s_2=\ldots=s_{|N(v)|}=0,\\
        1-s_v& otherwise.\\
    \end{cases}
\end{align}
As an example, for a vertex of degree 2, the resulting rank 3 tensor is
\begin{align}
    T(x_v)=\left(\begin{matrix}
    \left(\begin{matrix}
        0 &1 \\
        1 &1
    \end{matrix}\right)\\
    \left(\begin{matrix}
        x_v &0 \\
        0 &0
    \end{matrix}\right)
    \end{matrix}\right).
\end{align}

We do the same computation as independence polynomial, the coefficients of resulting polynomial gives the counting of maximal independent sets, or the maximal independence polynomial.
The computational complexity of this new tensor network is often larger than the one for computing independence polynomial.
However, in many sparse graphs, this tensor network contraction approach is still much faster than computing the maximal cliques on its complement by applying the Bron-Kerbosch algorithm.

@@

\section{Enumerating configurations}
One may also want to obtain all solutions, it can be achieved replacing the factors $a_x$ with a set of bit strings $s_x$,
We design a new element type having the following algebra
\begin{align}
\begin{split}
    s \oplus t &= s \cup t\\
    s \odot t &= \{\sigma \lor^\circ \tau | \sigma \in s, \tau \in t\}\\
    \mymathbb{0} &= \{\}\\
    \mymathbb{1} &= \{0^{\otimes n}\}
\end{split}\label{eq:set}
\end{align}
\iffalse
\begin{align}
\begin{split}
    s_x\infty^x \oplus s_y\infty^y &= \begin{cases}
        (s_x \cup s_y)\infty^{\max(x,y)}, & x = y\\
        s_y\infty^{\max(x,y)}, & x < y\\
        s_x\infty^{\max(x,y)}, & x > y
    \end{cases},\\
    s_x\infty^x \odot s_y\infty^y &= \{\sigma \lor^\circ \tau | \sigma \in s_x, \tau \in s_y\}\infty^{x+y},\\
    \mymathbb{0} &= \{\}\infty^{-\infty},\\
    \mymathbb{1} &= \{0^{\otimes n}\}\infty^{0},
\end{split}
\end{align}
\fi
where $\lor^\circ$ is the Hadamard logic or operation over two bit strings, which means joining of two local configurations.
The variable $x$ in the vertex tensor is initialized to $x_i = \{\boldsymbol{e}_{i}\}$,
where $\boldsymbol{e}_i$ is a one hot vector of size $|G|$.
One can easily check this algebra is a commutative semiring.
When we use the above algebra as factors of independence polynomials, the resulting algebra is also a commutative semiring.
With this new element type, the tensor network contraction will give all solutions rather than just a number for counting.
By slightly modifying the above algebra, it can also be used to obtain just a single configuration to save the computational effort.
%We leave this as an exercise for readers.

@@

\subsection{Bounding the enumeration space}
When one uses the set algebra in \Eq{eq:set} to represent the factors in \Eq{eq:countingtropical} for enumerating all optimum configurations, he will find the program stores more than necessary intermediate configurations and cause significant overheads in space.
To speed up the computation, we use $\alpha(G)$ to bound the searching space.
We first compute the value of $\alpha(G)$ with tropical numbers and cache all intermediate tensors.
Then we compute a boolean masks for each cached tensor, where we use a boolean true to represent a tensor element having contribution to the maximum independent set (i.e. with a non-zero gradient) and boolean false otherwise.
Finally, we perform masked matrix multiplication using the new element type with the above algebra for obtaining all configurations.
Notice that these masks are in fact tensor elements with non-zero gradients with respect to MIS size, we compute these masks by back propagating gradients.
To derive the backward rule, we consider a tropical matrix multiplication $C = A B$, we have the following inequality
\begin{align}
    A_{ij} \odot B_{jk} &\leq C_{ik}.
\end{align}
Moving $B_{ik}$ to the right hand side, we have
\begin{align}
    A_{ij} &\leq (\oplus_{k} (C_{ik}^{-1} \odot B_{jk}))^{-1}
\end{align}
where the tropical multiplicative inverse is defined as the additive inverse of the regular algebra. The equality holds if and only if element $A_{ij}$ contributions to $C$ (i.e. has non-zero gradient).
Let the mask for $C$ being $\overline C$, the backward rule for ``gradient'' masks reads
\begin{align}
\overline{A}_{ij} = \delta(A_{ij}, ((C^{\circ-1} \circ \overline C )B^T)_{ij}^{\circ -1}),
\end{align}
where ${}^{\circ -1}$ is the Hadamard inverse, $\circ$ is the Hadamard product, boolean false is treated as tropical zero and boolean true is treated as tropical one.
This rule defined on matrix multiplication can be easily generalized to tensor contraction by replacing the matrix multiplication between $C^{\circ-1} \circ \overline C$ and $B^T$ by a tensor contraction.~\cite{} \blue{maybe add an appendix?}  % my blog
